---
title: "Project 1 Code"
author: "Marshall Roll"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.width = 7, fig.height = 5, warning = FALSE, echo = FALSE)
```

```{r}
library(tidyverse)
library(naniar)
library(visdat)
library(table1)
library(gtsummary)
library(mice)
library(cmdstanr)
library(brms)
```

```{r}
# Raw data stored locally in Documents\Advanced_Methods_Roll\Project 1\Code
hiv_data_raw <- read_csv("../DataRaw/hiv_6624_final.csv")
```
# Filter Cohort
```{r}
# Select only complete cases
hiv_data <- hiv_data_raw %>% 
  filter(years %in% c(0, 2)) %>%
  group_by(newid) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  arrange(newid, years)

# Generate change in outcomes
hiv_outcomes <- hiv_data %>%
  # Apply log-10 transformation to viral load per investigator guidance 2/18/26
  mutate(VLOAD = log10(VLOAD)) %>% 
  pivot_wider(id_cols = newid,
              names_from = years,
              values_from = c(VLOAD, LEU3N, 
                              AGG_MENT, AGG_PHYS)) %>% 
  mutate(VLOAD_change = VLOAD_2 - VLOAD_0,
         LEU3N_change = LEU3N_2 - LEU3N_0, 
         PHYS_change = AGG_PHYS_2 - AGG_PHYS_0,
         MENT_change = AGG_MENT_2 - AGG_MENT_0) %>% 
  select(newid, VLOAD_change, LEU3N_change, PHYS_change,
         MENT_change, VLOAD_0, LEU3N_0, AGG_MENT_0,
         AGG_PHYS_0) %>% 
  rename(PHYS_0 = AGG_PHYS_0,
         MENT_0 = AGG_MENT_0)

# Create baseline covariates
hiv_covariates <- hiv_data %>% 
  mutate(race_collapsed = case_when(
    RACE %in% c(1) ~ "White",
    RACE %in% c(2, 3, 4, 5, 6, 7, 8) ~ "Non-white"
  ),
  educ_collapsed = case_when(
    EDUCBAS %in% c(1, 2, 3, 4) ~ "LessCollege",
    EDUCBAS %in% c(5, 6, 7) ~ "CollegeOrMore"
  ),
  smoking_status = case_when(
    SMOKE %in% c(1, 2) ~ "Non-smoker",
    SMOKE == 3 ~ "Smoker"
  ),
  # Removes missing values, coded as -1 and 999, and one value   newid = 179, with BMI ~514 (implausibly large)
  BMI = ifelse(BMI > 500 | BMI < 1, NA, BMI)) %>% 
  filter(years == 0) %>% 
  select(newid, age, BMI, smoking_status, race_collapsed,
         educ_collapsed, hard_drugs)

# Add 2-year adherence to covariates df
hiv_covariates <- hiv_covariates %>% 
  left_join(hiv_data %>% 
              filter(years == 2) %>% 
              select(newid, ADH), by = "newid")

hiv_cohort <- hiv_covariates %>% 
  left_join(hiv_outcomes, by = "newid")

# Only include cases with all outcomes non-missing in final cohort
hiv_cohort <- hiv_cohort %>% 
  filter(!is.na(VLOAD_change) & !is.na(LEU3N_change) &
         !is.na(PHYS_change) & !is.na(MENT_change))

# Relevel
hiv_cohort <- hiv_cohort %>% 
  mutate(educ_collapsed = 
           factor(educ_collapsed, 
                  levels = c("LessCollege", 
                             "CollegeOrMore")),
         race_collapsed = 
           factor(race_collapsed, 
                  levels = c("White", "Non-white")),
         adh_collapsed = case_when(
           ADH %in% c(1, 2) ~ TRUE,
           ADH %in% c(3, 4) ~ FALSE
         ))
```



# Assess Missingness

```{r}
# Little's MCAR Test
# p < 0.05, not MCAR
mcar_test <- mcar_test(hiv_data_raw %>% select(VLOAD, LEU3N, AGG_MENT, AGG_PHYS))
print(mcar_test)

# Function to find median value of each of the three outcomes in terms of missing values of fourth outcome
test_mnar <- function(variable, others){
  hiv_data_raw %>%
    filter(years %in% c(0, 2)) %>% 
    group_by(newid) %>% 
    mutate(max_year = max(years)) %>% 
    ungroup() %>% 
    bind_shadow() %>% 
    filter(years == 2) %>% 
    mutate(miss_2yr = ifelse(max_year == 0 | is.na({{ variable }}), 1, 0)) %>% 
    group_by(miss_2yr) %>% 
    summarise(across(all_of(others), 
                     list(median = ~median(.x, na.rm = TRUE),
                          n = ~n()
                          )))
}

test_mnar(VLOAD, c("LEU3N", "AGG_PHYS", "AGG_MENT"))
test_mnar(LEU3N, c("VLOAD", "AGG_PHYS", "AGG_MENT"))
test_mnar(AGG_PHYS, c("VLOAD", "LEU3N", "AGG_MENT"))
test_mnar(AGG_MENT, c("VLOAD", "LEU3N", "AGG_PHYS"))

# See medians are pretty similar across missingness status, little concern for harmful effect of MNAR 
```


# Table 1
```{r}
hiv_cohort %>% 
  mutate(
    hard_drugs = case_when(
      hard_drugs == 1 ~ "Hard Drug Use",
      hard_drugs == 0 ~ "No Hard Drug Use"
    ),
    educ_collapsed = case_when(
      educ_collapsed == "LessCollege" ~ "Less than College",
      educ_collapsed == "CollegeOrMore" ~ "College or More"
    ),
    race_collapsed = case_when(
      race_collapsed == "White" ~ "White, Non-Hispanic",
      race_collapsed == "Non-white" ~ "Other"
    )
  ) %>%
  mutate(educ_collapsed = 
           factor(educ_collapsed, 
                  levels = c("Less than College", 
                             "College or More")),
         race_collapsed = factor(race_collapsed, levels = c("White, Non-Hispanic", "Other"))) %>% 
  tbl_summary(
  by = hard_drugs, 
  include = c(age, BMI, smoking_status, race_collapsed,
         educ_collapsed, VLOAD_0, LEU3N_0, MENT_0,
         PHYS_0),
   statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    label = list(
      MENT_0 = "SF-36 MCS",
      PHYS_0 = "SF-36 PCS",
      age = "Age",
      smoking_status = "Smoking Status",
      race_collapsed = "Race",
      educ_collapsed = "Education",
      VLOAD_0 = "log10 Viral Load",
      LEU3N_0 = "CD4+ T-cell count"),
    missing_text = "(Missing)"
  )  %>% 
  add_overall()

# Be sure to note all covariate values at baseline
```

# Frequentist Regression

```{r}
# Function to run frequentist linear regression for 2-year change in each outcome, controlling for baseline level of outcome
run_freq_lm <- function(outcome) {
  
  formula_str <- paste0(outcome, "_change ~ hard_drugs + ", outcome, "_0 + age + BMI + smoking_status + race_collapsed + educ_collapsed")
  
  freq_mod <- lm(as.formula(formula_str), data = hiv_cohort)
  obj_name <- paste0(outcome, "_model")
  assign(obj_name, freq_mod, envir = .GlobalEnv)
  
  cat("\n--- Analysis for:", outcome, "---\n")
  
  print(summary(freq_mod))
}

run_freq_lm("VLOAD")
run_freq_lm("LEU3N")
run_freq_lm("PHYS")
run_freq_lm("MENT")
```
```{r}
set.seed(515)
run_bayes_lm <- function(outcome) {
  
 formula_str <- paste0(outcome, "_change ~ hard_drugs + ", outcome, "_0 + age + BMI + smoking_status + race_collapsed + educ_collapsed")
  
  bayes_mod <- brm(
    formula = as.formula(formula_str),
    data = hiv_cohort,
    family = gaussian(),
    prior = c(set_prior("normal(0,10000)", class='b'),
              set_prior("normal(0,10000)", class='Intercept')),
    backend = "cmdstanr",
    cores = 4,              
    chains = 4,             
    iter = 1000
  )
  
  obj_name <- paste0(outcome, "_bayes_model")
  assign(obj_name, bayes_mod, envir = .GlobalEnv)
  
  cat("\n--- Bayesian Analysis for:", outcome, "---\n")
  print(summary(bayes_mod))
  
  print(plot(bayes_mod)) 
}

run_bayes_lm("VLOAD")
run_bayes_lm("LEU3N")
run_bayes_lm("PHYS")
run_bayes_lm("MENT")
```


--Race: white, non-white
--No ethnicity
--Some college, College+
--Only use population with all outcomes non-missing
--log 10 viral load outcome
--keep track of sample size for each analysis
--data management can be supplementary but needs each step detailed correctly
--CAN have supplementary figures & tables
--missing values excluded from denominator of table 1 in each step
--Table 1 with only 506 in complete analysis
--Include sample size and number of observations in each figure caption
--no interactions
--0.5 log10 viral load, 2pts on QOL scores, 50 for CD4+ T cell counts: clinically meaningful changes
--adherence: 95% +, lower than 95%. 
--hard drug use -> adherence (descriptive)
--mediation analysis must be bayesian framework
--change in hard drug use between 
1) outcome ~ hard drug + cov
2) outcome ~ hard drug + ADH + cov

<!-- Raw data stored locally in Documents\Advanced_Methods_Roll\Project 1\Code -->

```{r}
sessionInfo()
```

