---
title: "Project 1 Code"
author: "Marshall Roll"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.width = 7, fig.height = 5, warning = FALSE, echo = FALSE)
```

```{r}
library(tidyverse)
library(naniar)
library(visdat)
library(table1)
library(gtsummary)
library(mice)
```

```{r}
# Raw data stored locally in Documents\Advanced_Methods_Roll\Project 1\Code
hiv_data_raw <- read_csv("../DataRaw/hiv_6624_final.csv")
```
# Filter Cohort
```{r}
# Select only complete cases
hiv_data <- hiv_data_raw %>% 
  filter(years %in% c(0, 2)) %>%
  group_by(newid) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  arrange(newid, years)

# Generate change in outcomes
hiv_outcomes <- hiv_data %>%
  # Apply log-10 transformation to viral load per investigator guidance 2/18/26
  mutate(VLOAD = log10(VLOAD)) %>% 
  pivot_wider(id_cols = newid,
              names_from = years,
              values_from = c(VLOAD, LEU3N, 
                              AGG_MENT, AGG_PHYS)) %>% 
  mutate(VLOAD_change = VLOAD_2 - VLOAD_0,
         LEU3N_change = LEU3N_2 - LEU3N_0, 
         PHYS_change = AGG_PHYS_2 - AGG_PHYS_0,
         MENT_change = AGG_MENT_2 - AGG_MENT_0) %>% 
  select(newid, VLOAD_change, LEU3N_change, PHYS_change,
         MENT_change, VLOAD_0, LEU3N_0, AGG_MENT_0,
         AGG_PHYS_0)

# Create baseline covariates
hiv_covariates <- hiv_data %>% 
  mutate(race_collapsed = case_when(
    RACE %in% c(1) ~ "White",
    RACE %in% c(2, 3, 4, 5, 6, 7, 8) ~ "Non-white"
  ),
  educ_collapsed = case_when(
    EDUCBAS %in% c(1, 2, 3, 4) ~ "LessCollege",
    EDUCBAS %in% c(5, 6, 7) ~ "CollegeOrMore"
  ),
  smoking_status = case_when(
    SMOKE %in% c(1, 2) ~ "Non-smoker",
    SMOKE == 3 ~ "Smoker"
  ),
  # Removes missing values, coded as -1 and 999, and one value   newid = 179, with BMI ~514 (implausibly large)
  BMI = ifelse(BMI > 500 | BMI < 1, NA, BMI)) %>% 
  filter(years == 0) %>% 
  select(newid, age, BMI, smoking_status, race_collapsed,
         educ_collapsed)

# Add 2-year adherence to covariates df
hiv_covariates <- hiv_covariates %>% 
  left_join(hiv_data %>% 
              filter(years == 2) %>% 
              select(newid, ADH), by = "newid")

hiv_cohort <- hiv_covariates %>% 
  left_join(hiv_outcomes, by = "newid")

# Only include cases with all outcomes non-missing in final cohort
hiv_cohort <- hiv_cohort %>% 
  filter(!is.na(VLOAD_change) & !is.na(LEU3N_change) &
         !is.na(PHYS_change) & !is.na(MENT_change))

by_subj <- hiv_cohort %>% 
  group_by(newid) %>% 
  mutate(avg_agg_ment = mean(AGG_MENT), 
         avg_agg_phys = mean(AGG_PHYS),
         avg_VLOAD = mean(VLOAD),
         avg_LEU3N = mean(LEU3N)) %>% 
  slice_head(n = 1)
```



# Assess Missingness
```{r}
gg_miss_var(hiv_data %>% 
              select(-c(LDL, TRIG, TCHOL, ADH, BMI)), 
            facet = any_hard_drugs)

gg_miss_var(hiv_data %>% 
              select(VLOAD, LEU3N, AGG_MENT, AGG_PHYS, ADH), 
            facet = ADH)

gg_miss_var(hiv_data %>% select(VLOAD, LEU3N, AGG_MENT, 
                                AGG_PHYS, years), 
            facet = years)

hiv_data %>%
  bind_shadow() %>%
  group_by(AGG_PHYS_NA) %>%
  summarise_at(.vars = c("VLOAD", "LEU3N", "AGG_MENT"),
               .funs = c("median", "sd", "min", "max"),
               na.rm = TRUE)

```

```{r}
# Little's MCAR Test
# p < 0.05, not MCAR
mcar_test <- mcar_test(hiv_data %>% select(VLOAD, LEU3N, AGG_MENT, AGG_PHYS))
print(mcar_test)


test_mnar <- function(data, outcome_vars) {
  results <- list()
  
  for (outcome in outcome_vars) {
    # Create missingness indicator for each outcome
    miss_indicator <- paste0(outcome, "_NA")
    
    # Test if missingness is associated with OTHER observed outcomes
    other_outcomes <- setdiff(outcome_vars, outcome)
    
    # Logistic regression: does missingness in outcome Y depend on observed values of other three outcomes?
    for (predictor in other_outcomes) {
      formula_str <- paste0("as.factor(", miss_indicator, ") ~ ", predictor)
      
      tryCatch({
        model <- glm(as.formula(formula_str), 
                     data = data %>% bind_shadow(), 
                     family = binomial)
        
        results[[paste(outcome, "missing_by", predictor)]] <- broom::tidy(model) %>%
          filter(term == predictor) %>%
          mutate(outcome_missing = outcome,
                 predictor_observed = predictor)
      }, error = function(e) {
        message(paste("Error fitting model for", outcome, "~", predictor))
      })
    }
  }
  
  bind_rows(results)
}

outcome_vars <- c("VLOAD", "LEU3N", "AGG_MENT", "AGG_PHYS")
mnar_results <- test_mnar(hiv_data, outcome_vars)

mnar_results %>%
  select(outcome_missing, predictor_observed, estimate, p.value) %>%
  arrange(p.value)

```


# Table 1
```{r}
hiv_data %>% 
  group_by(newid) %>% 
  tbl_summary(
  by = any_hard_drugs, 
  include = c(AGG_MENT),
   statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    label = list(AGG_MENT = "SF-36 Mental Component Summary (MCS)"),
    missing_text = "(Missing)"
  ) 
```

--Race: white, non-white
--No ethnicity
--Some college, College+
--Only use population with all outcomes non-missing
--log 10 viral load outcome
--keep track of sample size for each analysis
--data management can be supplementary but needs each step detailed correctly
--CAN have supplementary figures & tables
--missing values excluded from denominator of table 1 in each step
--Table 1 with only 506 in complete analysis
--Include sample size and number of observations in each figure caption
--no interactions
--0.5 log10 viral load, 2pts on QOL scores, 50 for CD4+ T cell counts: clinically meaningful changes

<!-- Raw data stored locally in Documents\Advanced_Methods_Roll\Project 1\Code -->
