---
title: "Project 1 Code"
author: "Marshall Roll"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.width = 7, fig.height = 5, warning = FALSE, echo = FALSE)
```

```{r}
library(tidyverse)
library(naniar)
library(visdat)
library(table1)
library(gtsummary)
library(mice)
library(cmdstanr)
library(brms)
library(gt)
library(flextable)
```

```{r}
# Raw data stored locally in Documents\Advanced_Methods_Roll\Project 1\Code
hiv_data_raw <- read_csv("../DataRaw/hiv_6624_final.csv")
```

# Filter Cohort
```{r}
# Select only complete cases
hiv_data <- hiv_data_raw %>% 
  filter(years %in% c(0, 2)) %>%
  group_by(newid) %>%
  filter(n() == 2) %>%
  ungroup() %>%
  arrange(newid, years)

# Generate change in outcomes
hiv_outcomes <- hiv_data %>%
  # Apply log-10 transformation to viral load per investigator guidance 2/18/26
  mutate(VLOAD = log10(VLOAD)) %>% 
  pivot_wider(id_cols = newid,
              names_from = years,
              values_from = c(VLOAD, LEU3N, 
                              AGG_MENT, AGG_PHYS)) %>% 
  mutate(VLOAD_change = VLOAD_2 - VLOAD_0,
         LEU3N_change = LEU3N_2 - LEU3N_0, 
         PHYS_change = AGG_PHYS_2 - AGG_PHYS_0,
         MENT_change = AGG_MENT_2 - AGG_MENT_0) %>% 
  select(newid, VLOAD_change, LEU3N_change, PHYS_change,
         MENT_change, VLOAD_0, LEU3N_0, AGG_MENT_0,
         AGG_PHYS_0) %>% 
  rename(PHYS_0 = AGG_PHYS_0,
         MENT_0 = AGG_MENT_0)

# Create baseline covariates
hiv_covariates <- hiv_data %>% 
  mutate(race_collapsed = case_when(
    RACE %in% c(1) ~ "White",
    RACE %in% c(2, 3, 4, 5, 6, 7, 8) ~ "Non-white"
  ),
  educ_collapsed = case_when(
    EDUCBAS %in% c(1, 2, 3, 4) ~ "LessCollege",
    EDUCBAS %in% c(5, 6, 7) ~ "CollegeOrMore"
  ),
  smoking_status = case_when(
    SMOKE %in% c(1, 2) ~ "Non-smoker",
    SMOKE == 3 ~ "Smoker"
  ),
  # Removes missing values, coded as -1 and 999, and one value   newid = 179, with BMI ~514 (implausibly large)
  BMI = ifelse(BMI > 500 | BMI < 1, NA, BMI)) %>% 
  filter(years == 0) %>% 
  select(newid, age, BMI, smoking_status, race_collapsed,
         educ_collapsed, hard_drugs)

# Add 2-year adherence to covariates df
hiv_covariates <- hiv_covariates %>% 
  left_join(hiv_data %>% 
              filter(years == 2) %>% 
              select(newid, ADH), by = "newid")

hiv_cohort <- hiv_covariates %>% 
  left_join(hiv_outcomes, by = "newid")

# Only include cases with all outcomes non-missing in final cohort
hiv_cohort <- hiv_cohort %>% 
  filter(!is.na(VLOAD_change) & !is.na(LEU3N_change) &
         !is.na(PHYS_change) & !is.na(MENT_change))

# Relevel
hiv_cohort <- hiv_cohort %>% 
  mutate(educ_collapsed = 
           factor(educ_collapsed, 
                  levels = c("LessCollege", 
                             "CollegeOrMore")),
         race_collapsed = 
           factor(race_collapsed, 
                  levels = c("White", "Non-white")),
         adh_collapsed = case_when(
           ADH %in% c(1, 2) ~ 1,
           ADH %in% c(3, 4) ~ 0
         ))

hiv_cohort <- hiv_cohort %>% 
  select(-ADH) %>%
  drop_na(BMI)
```

# Assess Missingness

```{r}
# Little's MCAR Test
# p < 0.05, not MCAR
mcar_test <- mcar_test(hiv_data_raw %>% select(VLOAD, LEU3N, AGG_MENT, AGG_PHYS))
print(mcar_test)

# Function to find median value of each of the three outcomes in terms of missing values of fourth outcome  
test_mnar <- function(variable, others){
  hiv_data_raw %>%
    filter(years %in% c(0, 2)) %>% 
    group_by(newid) %>% 
    mutate(max_year = max(years)) %>% 
    ungroup() %>% 
    bind_shadow() %>% 
    filter(years == 2) %>% 
    mutate(miss_2yr = ifelse(max_year == 0 | is.na({{ variable }}), 1, 0)) %>% 
    group_by(miss_2yr) %>% 
    summarise(across(all_of(others), 
                     list(median = ~median(.x, na.rm = TRUE),
                          n = ~n()
                          )))
}

test_mnar(VLOAD, c("LEU3N", "AGG_PHYS", "AGG_MENT"))
test_mnar(LEU3N, c("VLOAD", "AGG_PHYS", "AGG_MENT"))
test_mnar(AGG_PHYS, c("VLOAD", "LEU3N", "AGG_MENT"))
test_mnar(AGG_MENT, c("VLOAD", "LEU3N", "AGG_PHYS"))

# See medians are pretty similar across missingness status, little concern for harmful effect of MNAR 
```


# Table 1
```{r}
hiv_cohort %>% 
  mutate(
    hard_drugs = case_when(
      hard_drugs == 1 ~ "Hard Drug Use",
      hard_drugs == 0 ~ "No Hard Drug Use"
    ),
    educ_collapsed = case_when(
      educ_collapsed == "LessCollege" ~ "Less than College",
      educ_collapsed == "CollegeOrMore" ~ "College or More"
    ),
    race_collapsed = case_when(
      race_collapsed == "White" ~ "White, Non-Hispanic",
      race_collapsed == "Non-white" ~ "Other"
    ),
    adh_collapsed = case_when(
      adh_collapsed == 1 ~ "≥95% Adherence",
      adh_collapsed == 0 ~ "<95 % Adherence"
    )
  ) %>%
  mutate(educ_collapsed = 
           factor(educ_collapsed, 
                  levels = c("Less than College", 
                             "College or More")),
         race_collapsed = factor(race_collapsed, levels = c("White, Non-Hispanic", "Other")),
         adh_collapsed = factor(adh_collapsed, levels = c("≥95% Adherence", "<95 % Adherence"))) %>% 
  tbl_summary(
  by = hard_drugs, 
  include = c(age, BMI, smoking_status, race_collapsed,
         educ_collapsed, VLOAD_0, LEU3N_0, MENT_0,
         PHYS_0, adh_collapsed),
   statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    label = list(
      MENT_0 = "SF-36 MCS",
      PHYS_0 = "SF-36 PCS",
      age = "Age",
      smoking_status = "Smoking Status",
      race_collapsed = "Race",
      educ_collapsed = "Education",
      VLOAD_0 = "log10 Viral Load",
      LEU3N_0 = "CD4+ T-cell count",
      adh_collapsed = "2-year HAART Adherence"),
    missing_text = "(Missing)"
  )  %>% 
  add_overall()

# Be sure to note all covariate values at baseline
```

# Frequentist Regression

```{r}
# Function to run frequentist linear regression for 2-year change in each outcome, controlling for baseline level of outcome
run_freq_lm <- function(outcome) {
  
  formula_str <- paste0(outcome, "_change ~ hard_drugs + ", outcome, "_0 + age + BMI + smoking_status + race_collapsed + educ_collapsed")
  
  freq_mod <- lm(as.formula(formula_str), data = hiv_cohort)
  obj_name <- paste0(outcome, "_model")
  assign(obj_name, freq_mod, envir = .GlobalEnv)
  
  cat("\n--- Analysis for:", outcome, "---\n")
  
  print(summary(freq_mod))
}

run_freq_lm("VLOAD")
run_freq_lm("LEU3N")
run_freq_lm("PHYS")
run_freq_lm("MENT")
```
```{r}
set.seed(515)
run_bayes_lm <- function(outcome) {
  
 formula_str <- paste0(outcome, "_change ~ hard_drugs + ", outcome, "_0 + age + BMI + smoking_status + race_collapsed + educ_collapsed")
  
  bayes_mod <- brm(
    formula = as.formula(formula_str),
    data = hiv_cohort,
    family = gaussian(),
    prior = c(set_prior("normal(0,100)", class='Intercept'),
              set_prior("normal(0,100)", class='b'),
              set_prior("normal(0,100)", class='sigma') # BRMS will automatically convert this to half-normal
              ),
    backend = "cmdstanr",
    cores = 4,              
    chains = 4,             
    iter = 2000,
    warmup = 500
  )
  
  obj_name <- paste0(outcome, "_bayes_model")
  assign(obj_name, bayes_mod, envir = .GlobalEnv)
  
  cat("\n--- Bayesian Analysis for:", outcome, "---\n")
  print(summary(bayes_mod))
  
  # print(plot(bayes_mod)) 
}

run_bayes_lm("VLOAD")
run_bayes_lm("LEU3N")
run_bayes_lm("PHYS")
run_bayes_lm("MENT")
```

# Mediation Analysis
<!-- 
Hypothesis: Adherence may serve as a mediator or confounder between hard drug use and outcomes. Need to establish association between HARD_DRUG -> ADH and ADH -> outcomes. Hard drugs and adherence will be assessed using descriptive statistics (both binary) and models adjusted for adherence will also be fit for each outcome. No formal mediation analysis will be conducted, rather the difference in hard drug use coefficients will be presented between adjusted vs unadjusted models. 
-->

# Descriptive Analysis
```{r}
fisher.test(hiv_cohort$hard_drugs, hiv_cohort$adh_collapsed)
```

# Frequentist Adherence Models
```{r}
# Function to run frequentist linear regression for 2-year change in each outcome with adherence
run_freq_lm_adh <- function(outcome) {
  
  formula_str <- paste0(outcome, "_change ~ hard_drugs + ", outcome, "_0 + adh_collapsed + age + BMI + smoking_status + race_collapsed + educ_collapsed")
  
  freq_mod <- lm(as.formula(formula_str), data = hiv_cohort)
  obj_name <- paste0(outcome, "_adh_model")
  assign(obj_name, freq_mod, envir = .GlobalEnv)
  
  cat("\n--- Analysis for:", outcome, "---\n")
  
  print(summary(freq_mod))
}

run_freq_lm_adh("VLOAD")
run_freq_lm_adh("LEU3N")
run_freq_lm_adh("PHYS")
run_freq_lm_adh("MENT")
```

# Bayesian Adherence Models
```{r}
set.seed(515)
run_bayes_adh_lm <- function(outcome) {
  
 formula_str <- paste0(outcome, "_change ~ hard_drugs + ", outcome, "_0 + adh_collapsed + age + BMI + smoking_status + race_collapsed + educ_collapsed")
  
  bayes_mod_adh <- brm(
    formula = as.formula(formula_str),
    data = hiv_cohort,
    family = gaussian(),
    prior = c(set_prior("normal(0,100)", class='Intercept'),
              set_prior("normal(0,100)", class='b'),
              set_prior("normal(0,100)", class='sigma')  # BRMS will automatically convert this to half-normal
              ),
    backend = "cmdstanr",
    cores = 4,              
    chains = 4,             
    iter = 2000,
    warmup = 500
  )
  
  obj_name_adh <- paste0(outcome, "_adh_bayes_model")
  assign(obj_name_adh, bayes_mod_adh, envir = .GlobalEnv)
  
  cat("\n--- Bayesian Analysis with ADH for:", outcome, "---\n")
  print(summary(bayes_mod_adh))
  
  # print(plot(bayes_mod)) 
}

run_bayes_adh_lm("VLOAD")
run_bayes_adh_lm("LEU3N")
run_bayes_adh_lm("PHYS")
run_bayes_adh_lm("MENT")
```
<!-- Good mixing, everything normally distributed with equal tails. -->

# Results Tables

```{r}
# Helper function to extract freq model results
extract_freq <- function(model, predictor = "hard_drugs") {
  coef_table <- summary(model)$coefficients
  est <- coef_table[predictor, "Estimate"]
  se  <- coef_table[predictor, "Std. Error"]
  pval <- coef_table[predictor, "Pr(>|t|)"] 
  pval <- round(pval, 3)
  ci_lo <- est - 1.96 * se
  ci_hi <- est + 1.96 * se
  list(
    coef_ci = sprintf("%.3f (%.3f, %.3f)", est, ci_lo, ci_hi),
    pval = ifelse(pval == 0.000, "<0.001", pval)
  )
}

# Helper function to extract Bayesian model results
extract_bayes <- function(model, predictor = "hard_drugs") {
  effects <- fixef(model)
  row_number <- which(rownames(effects) == "hard_drugs")
  est <- effects[row_number, 1]
  ci_lo <- effects[row_number ,3]
  ci_hi <- effects[row_number ,4]
  waic <- waic(model)$estimates["waic", "Estimate"]
  list(
    coef_ci = sprintf("%.3f (%.3f, %.3f)", est, ci_lo, ci_hi),
    waic = sprintf("%.1f", waic)
  )
}

outcomes <- list(
  list(name = "log10 Viral Load", 
       freq = VLOAD_model,      
       freq_adh = VLOAD_adh_model,
       bayes = VLOAD_bayes_model,
       bayes_adh = VLOAD_adh_bayes_model),
  list(name = "CD4+ T-cell Count",  
       freq = LEU3N_model,
       freq_adh = LEU3N_adh_model,
       bayes = LEU3N_bayes_model, 
       bayes_adh = LEU3N_adh_bayes_model),
  list(name = "SF-36 PHYS Score",
       freq = PHYS_model,
       freq_adh = PHYS_adh_model,
       bayes = PHYS_bayes_model,  
       bayes_adh = PHYS_adh_bayes_model),
  list(name = "SF-36 MENT Score", 
       freq = MENT_model,
       freq_adh = MENT_adh_model,
       bayes = MENT_bayes_model,  
       bayes_adh = MENT_adh_bayes_model)
)

rows <- list()
for (o in outcomes) {
  rows[[length(rows) + 1]] <- data.frame(
    Outcome = o$name, Model = "", Freq_CoefCI = "", Pval = "", Bayes_CoefCI = "", WAIC = "",
    stringsAsFactors = FALSE
  )
  # Unadjusted
  f1 <- extract_freq(o$freq)
  b1 <- extract_bayes(o$bayes)
  rows[[length(rows) + 1]] <- data.frame(
    Outcome = "", Model = "Unadjusted",
    Freq_CoefCI = f1$coef_ci, Pval = as.character(f1$pval),
    Bayes_CoefCI = b1$coef_ci, WAIC = b1$waic,
    stringsAsFactors = FALSE
  )
  # Adjusted
  f2 <- extract_freq(o$freq_adh)
  b2 <- extract_bayes(o$bayes_adh)
  rows[[length(rows) + 1]] <- data.frame(
    Outcome = "", Model = "Adjusted",
    Freq_CoefCI = f2$coef_ci, Pval = as.character(f2$pval),
    Bayes_CoefCI = b2$coef_ci, WAIC = b2$waic,
    stringsAsFactors = FALSE
  )
}

tbl <- bind_rows(rows)

tbl %>% 
  gt() %>% 
  cols_label(
    Outcome = "Outcome",
    Model = "Model",
    Freq_CoefCI = "Hard Drug Use Effect (95% CI)",
    Pval = "p-value",
    Bayes_CoefCI = "Hard Drug Use Effect (95% CrI)",
    WAIC = "WAIC"
  ) %>% 
  tab_spanner(label = "Frequentist", columns = c(Freq_CoefCI, Pval)) %>% 
  tab_spanner(label = "Bayesian", columns = c(Bayes_CoefCI, WAIC)) %>% 
  tab_style(
    style = cell_text(),
    locations = cells_body(columns = Outcome, rows = Outcome != "")
  ) %>% 
  tab_footnote("Adjusted models include adherence as a covariate in addition to the standard covariates in the unadjusted model. Effect of hard drug use represents the difference in predicted 2-year change in outcomes of an average hard drug user relative to an average non-hard drug user. 95% CI = frequentist confidence interval; 95% CrI = Bayesian credible interval.") %>%
  tab_options(table.font.size = 12)
```

# Figures

```{r}
extract_bayes_std_coef <- function(model, data = hiv_cohort, outcome, predictor = "b_hard_drugs") {
  outcome_sd <- sd(data[[outcome]], na.rm = TRUE)
  effects <- as_draws_df(model)
  draws <- effects[[predictor]] / outcome_sd
  data.frame(
    est = mean(draws),
    lo = quantile(draws, 0.025),
    hi = quantile(draws, 0.975)
  )
}

plot_results <- bind_rows(
  extract_bayes_std_coef(VLOAD_bayes_model, outcome = "VLOAD_change") %>%
    mutate(outcome = "log10 Viral Load"),
  extract_bayes_std_coef(LEU3N_bayes_model, outcome = "LEU3N_change") %>%
    mutate(outcome = "CD4+ T-cell count"),
  extract_bayes_std_coef(PHYS_bayes_model, outcome = "PHYS_change") %>%
    mutate(outcome = "SF-36 PCS"),
  extract_bayes_std_coef(MENT_bayes_model, outcome = "MENT_change") %>%
    mutate(outcome = "SF-36 MCS")
) %>%
  mutate(outcome = factor(outcome, levels = c("SF-36 MCS", "SF-36 PCS",  "CD4+ T-cell count", "log10 Viral Load")))

ggplot(plot_results, aes(x = est, y = outcome)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = lo, xmax = hi),
                 height = 0.15, linewidth = 0.7) +
  geom_point(size = 3, color = "darkcyan") +
  labs(
    x = "Standardized coefficient (SD units)",
    y = "",
    title = "Effect of Hard Drug Use on HIV Outcomes",
    caption = "Posterior means ± 95% credible intervals (standardized).  Effects measured as predicted difference\n in 2-year change in outcome between a typical hard drug user and a typical non-hard drug user.",
  ) +
  theme_bw(base_size = 13) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(hjust = 1)
  )
```



--Race: white, non-white
--No ethnicity
--Some college, College+
--Only use population with all outcomes non-missing
--log 10 viral load outcome
--keep track of sample size for each analysis
--data management can be supplementary but needs each step detailed correctly
--CAN have supplementary figures & tables
--missing values excluded from denominator of table 1 in each step
--Table 1 with only 506 in complete analysis
--Include sample size and number of observations in each figure caption
--no interactions
--0.5 log10 viral load, 2pts on QOL scores, 50 for CD4+ T cell counts: clinically meaningful changes
--adherence: 95% +, lower than 95%. 
--hard drug use -> adherence (descriptive)
--mediation analysis must be bayesian framework
--change in hard drug use between 
1) outcome ~ hard drug + cov
2) outcome ~ hard drug + ADH + cov


Raw data stored locally in `Documents\Advanced_Methods_Roll\Project 1\Code`.

```{r}
sessionInfo()
```
