---
title: "Project 0 Code"
author: "Marshall Roll"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(tidyverse)
library(rmcorr)
library(naniar)
library(lmerTest)
library(flextable)
library(ggridges)
```

```{r}
cort_data <- read_csv("../DataRaw/Project0_Clean_v2.csv")
```
# Cleaning

```{r}
# Rename variables and collect into working df
cort_final <- cort_data %>% mutate(
  date = `Collection Date`,
  sample = `Collection Sample`, 
  book_time = `Booket: Clock Time`,
  cap_time = `MEMs: Clock Time`,
  wake_time = `Sleep Diary reported wake time`,
  cortisol = `Cortisol (nmol/L)`,
  dhea = `DHEA (nmol/L)`
) %>% 
  select(SubjectID, date, sample, book_time, cap_time, wake_time, cortisol, dhea)

# Carry wake time across all observations for calculation
cort_final <- cort_final %>% 
  group_by(SubjectID, date) %>% 
  mutate(wake_time = replace_na(wake_time, first(wake_time))) %>% 
  ungroup()

# Minutes since wake time for cap & booklet
cort_final <- cort_final %>% 
  mutate(wake_to_cap = as.numeric(cap_time - wake_time)/60,
         wake_to_book = as.numeric(book_time - wake_time)/60) 

# Replace cortisol levels greater than 80 nmol/L per investigator guidance
cort_final <- cort_final %>% 
  mutate(cortisol = ifelse(cortisol > 80, NA, cortisol))
```

# RQ1 - Cap & Book Time Agreement
```{r}
# Repeated measures correlation - only takes into account complete cases
# Also only measures correlation within subjects -- does not alert to systematic biases
rep_corr <- rmcorr(participant = factor(SubjectID), 
                 measure1 = wake_to_book, 
                 measure2 = wake_to_cap,
                 dataset = cort_final)

rep_corr

# R = 0.99

# Subject-specific correlation measures visualization
cort_final %>% 
  ggplot(aes(x = wake_to_book, y = wake_to_cap, color = factor(SubjectID),
             group = factor(SubjectID))) + 
  geom_point() + 
  geom_line(stat = "smooth", method = 'lm', se = FALSE, alpha = 0.15, linewidth = 0.8) +
  scale_color_viridis_d(option = "H") + 
  theme_minimal() +
  guides(color = "none") + 
  labs(x = "Minutes Since Wake (Booklet)", y = "Minutes Since Wake (Cap)")
 
# Proportion of missingness across time points
# Between 20-30% for all time pts.
cort_final %>%
  group_by(sample) %>%
  summarise(
    n_total = n(),
    pct_complete = round(sum(!is.na(wake_to_cap) & !is.na(wake_to_book)) / n_total * 100, 1),
    pct_complete_cap = round(sum(!is.na(wake_to_cap))/n() * 100, 1),
    pct_complete_book = round(sum(!is.na(wake_to_book))/n() * 100, 1)
  ) %>% 
  flextable() %>% 
  set_header_labels(sample = "Sample",
                    n_total = "n",
                    pct_complete = "Complete (%)", 
                    pct_complete_cap = "Complete Cap (%)",
                    pct_complete_book = "Complete Book (%)")

# Mixed model for assessing correlation with random intercept for subject
# Shows a moderate ~10 min difference between average cap and book time from wake 
corr_glmm <- lmer(wake_to_cap ~ wake_to_book + (1 | SubjectID), 
              data = cort_final)

summary(corr_glmm)

# Pretty good agreement, going to use book time because there is less missingness going forward
```

To assess the agreement between cap and booklet time, a random intercepts model was fit with cap time since wake as the outcome and booklet time as the outcome. There is a slight systematic difference between the two recording methods, with cap readings occurring an average of about 10 minutes later than booklet recordings and slight subject-specific variability (random intercept SD = 6 minutes). Within-subject correlation for complete cases is very high (repeated measures correlation, r = 0.99). Between 20-30% of observations were incomplete at each of the four time points, with most time points having higher levels of missingness in the cap compared to the book.  

# RQ2: Adherence to Timing Schedule

```{r}
# Want to test adherence in second sample (how close to 30 mins) and final sample (how close to 600)
# Adequate: within 15 mins, Good: within 7.5 minutes

# Sample 2
cort_samp2 <- cort_final %>% filter(sample == 2)

cort_samp2 <- cort_samp2 %>% 
  mutate(within_7.5 = ifelse(wake_to_book > 23 & wake_to_book < 37, 1, 0),
         within_15 = ifelse(wake_to_book > 15 & wake_to_book < 45, 1, 0))

cort_samp2 %>% summarize(
  prop_7.5 = round(sum(within_7.5, na.rm = TRUE)/n() * 100, 1),
  prop_15 = round(sum(within_15, na.rm = TRUE)/n() * 100, 1)
)

# Sample 4
cort_samp4 <- cort_final %>% filter(sample == 4)

cort_samp4 <- cort_samp4 %>% 
  mutate(within_7.5 = ifelse(wake_to_book > 593 & wake_to_book < 607, 1, 0),
         within_15 = ifelse(wake_to_book > 585 & wake_to_book < 615, 1, 0))

cort_samp4 %>% summarize(
  prop_7.5 = round(sum(within_7.5, na.rm = TRUE)/n() * 100, 1),
  prop_15 = round(sum(within_15, na.rm = TRUE)/n() * 100, 1)
)

# Adherence for 600 minutes is low (35.5% within 15 minutes)
# Density plot to assess full distribution

cort_final %>%
  filter(sample %in% c(2, 3, 4)) %>% 
  mutate(sample = factor(sample, 
                         levels = c(4, 3, 2),
                         labels = c("Sample 4 (Bedtime)", 
                                   "Sample 3 (5pm-8pm)",
                                   "Sample 2 (11am-2pm)"))) %>% 
  ggplot(aes(x = wake_to_book, y = sample, fill = sample)) + 
  geom_density_ridges(alpha = 0.7, scale = 0.9) +
  scale_fill_manual(values = c("#e6194B", "#f58231", "#4363d8")) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  ) +
  labs(x = "Minutes Since Wake (Booklet)", 
       y = NULL,
       title = "Distribution of Time Since Waking by Sample")
```

Outcome: time since reported wakeup time
Agreement: On average, are cap and booklet time similar? 
1) Test of association, rate of bias (sleep diary wake time - book time/Cap time; calculate manually)
2) Table/figs, adherence: good = 7.5 mins, adequate = 15 mins
3) Wakeup - 30, rate of decline post-30 mins, use moles

26+ cortisol possible but unlikely
exclude cortisol over 80

Upper Limit DHEA: 5.205 nmol/L

1) Mixed effects model with RI for subject or Repeated Measures correlation?  
 -- scatterplot for booklet time since wake vs cap time    
2) Tables/figures 
3) 2 random effects models 





