---
title: "Project 3 Analysis"
author: "Marshall Roll"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(tidyverse)
library(survival)
library(survminer)
library(glmnet)
```

```{r}
# Raw data stored locally in Documents\Advanced_Methods_Roll\Project 3\Code
heart <- read_csv("../DataRaw/frmgham2.csv")
```

# Data Cleaning
```{r}
# n = 4434
heart <- heart %>% 
  filter(PERIOD == 1) %>% #baseline covariates
  filter(PREVSTRK != 1) %>% #exclude prior stroke # n = 4402
  mutate(time_stroke_yrs = TIMESTRK/365.25) %>% # Convert time stroke to years
  mutate(
    time_stroke_yrs = pmin(time_stroke_yrs, 10),
    censored = ifelse(time_stroke_yrs < 10, 1, 0),
    SEX = case_when(SEX == 1 ~ "Male",
                     SEX == 2 ~ "Female")
  )

heart_male <- heart %>% filter(SEX == "Male") # n = 1930
heart_female <- heart %>% filter(SEX == "Female") # n = 2472
```

# Table 1
```{r}
heart %>% 
  mutate(
    hard_drugs = case_when(
      hard_drugs == 1 ~ "Hard Drug Use",
      hard_drugs == 0 ~ "No Hard Drug Use"
    ), ...
  ) %>%
  tbl_summary(
  by = hard_drugs, 
  include = c(age, BMI, smoking_status, race_collapsed,
         educ_collapsed, VLOAD_0, LEU3N_0, MENT_0,
         PHYS_0, adh_collapsed),
   statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    label = list(
      MENT_0 = "SF-36 MCS",
      PHYS_0 = "SF-36 PCS",
      age = "Age",
      smoking_status = "Smoking Status",
      race_collapsed = "Race",
      educ_collapsed = "Education",
      VLOAD_0 = "log10 Viral Load",
      LEU3N_0 = "CD4+ T-cell count",
      adh_collapsed = "2-year HAART Adherence"),
    missing_text = "(Missing)"
  )  %>% 
  add_overall()

```

# Overall

```{r}
surv<-survfit(Surv(time_stroke_yrs, censored)~SEX, data=heart)

ggsurvplot(surv, xlim=c(0, 10))
```

# Male

```{r}
coxph(Surv(time_stroke_yrs, censored)~AGE + DIABETES + SYSBP, 
      data = heart_male)
```
# Female

```{r}
coxph(Surv(time_stroke_yrs, censored)~AGE + DIABETES + SYSBP, 
      data = heart_female)
```


# Questions
-- Clarify time frame: first ten years of follow-up? Yes, 3 periods? Use 365.25 * 10 as 10 years, don't include period 3 (1st period only)
  -- time 0, period 1, stroke 1 = stroke before 
-- Competing risks (comment on limitations but don't account for them)
-- Censor strokes after 10 years
-- exclusion criteria: prior stroke
-- Risk profile: know want to look at age, diabetes, systolic blood pressure; not sure about CHD, blood pressure meds, smoking, total cholesterol, BMI: baseline only 
--Want to keep as much data as possible
--Risk profiles: estimates for 10-year survival for ages 40, 50, and 60 across baseline no risk factors (diabetes, SBP, maybe CHD, BP Meds, smoker, cholesterol, BMI: average continuous, 0's on categoricals including BP meds); high blood pressure (SYS 160+); diabetes; high blood pressure and diabetes; 1 additional profile based on included covariates  
--want survival prob for person of each profile
--More interested in within-sex differences; not necessarily interested in between-sex
--PH assumption: Look at for everything

only descriptive time varying on diabetes and sys BP
2 separate models for M/F (diff covariates, etc.)

Null: HR = 1 for each of age, diabetes, sys BP
367 competing deaths, 111 strokes; standard cox model treats deaths as censored, which overestimates cumulative stroke incidence 

Should be 32 excluded for stroke
Table 1
----
Stroke during follow up
Follow up Time

n = 4275 (127 missing)
no need to exclude high BP & cholesterol
