---
title: "Project 3 Analysis"
author: "Marshall Roll"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(tidyverse)
library(survival)
library(survminer)
library(glmnet)
library(gtsummary)
library(cowplot)
library(naniar)
```

```{r}
# Raw data stored locally in Documents\Advanced_Methods_Roll\Project 3\Code
heart <- read_csv("../DataRaw/frmgham2.csv")
```

# Data Cleaning
```{r}
# n = 4434
heart <- heart %>% 
  filter(PERIOD == 1) %>% #baseline covariates
  filter(PREVSTRK != 1) %>% #exclude prior stroke # n = 4402
  mutate(time_stroke_yrs = TIMESTRK/365.25) %>% # Convert time stroke to years
  mutate(
    time_stroke_yrs = pmin(time_stroke_yrs, 10),
    censored = ifelse(time_stroke_yrs < 10, 1, 0),
    SEX = case_when(SEX == 1 ~ "Male",
                     SEX == 2 ~ "Female")
  )

heart_male <- heart %>% filter(SEX == "Male") # n = 1930
heart_female <- heart %>% filter(SEX == "Female") # n = 2472
```

# Table 1
```{r}
heart %>% 
  tbl_summary(
    by = SEX, 
    include = c(AGE, SYSBP, DIABP, TOTCHOL, CURSMOKE, BMI, DIABETES, BPMEDS,
                PREVCHD, PREVMI, STROKE, HOSPMI, DEATH, time_stroke_yrs),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)" 
      ),
    digits = all_continuous() ~ 2,
    label = list(
      AGE = "Age",
      SYSBP = "Systolic BP",
      DIABP = "Diastolic BP",
      CURSMOKE = "Smoker",
      TOTCHOL = "Total Cholesterol (mg/dL)",
      DIABETES = "Diabetes",
      BPMEDS = "Blood Pressure Medication",
      PREVCHD = "Coronary Heart Disease", 
      PREVMI = "Past MI",
      STROKE = "10-year Stroke",
      HOSPMI = "10-year MI",
      DEATH = "10-year Death",
      time_stroke_yrs ~ "Follow-up Time (Years)"
    ),
    missing_text = "(Missing)"
  ) %>% 
  add_overall()

```

# Missing Data

```{r}
## No super high rates of stroke among missing data levels
heart %>% 
  group_by(is.na(BPMEDS), SEX) %>% 
  summarize(Stroke = sum(STROKE), prop_stroke = sum(STROKE)/n())

heart %>% 
  group_by(is.na(BMI), SEX) %>% 
  summarize(Stroke = sum(STROKE), prop_stroke = sum(STROKE)/n())

heart %>% 
  group_by(is.na(TOTCHOL), SEX) %>% 
  summarize(Stroke = sum(STROKE), prop_stroke = sum(STROKE)/n())

mcar_test(heart_male %>% select(BMI, TOTCHOL, BMI)) # p = 0.14, could be MCAR
```


# Fit Models

```{r}
inspect_lasso_fit <- function(df) {
  forced <- c("AGE", "DIABETES", "SYSBP")
  penalized <- c("PREVCHD", "BPMEDS", "CURSMOKE", "TOTCHOL", "BMI")
  all_vars <- c(forced, penalized)

  surv_obj <- Surv(df$time_stroke_yrs, df$censored)
  X <- df %>%
    select(all_of(all_vars)) %>%
    mutate(across(everything(), as.numeric)) %>%
    as.matrix()

  pf <- c(rep(0, length(forced)), rep(1, length(penalized)))

  set.seed(285)
  cv_fit <- cv.glmnet(
    x = X,
    y = surv_obj,
    family = "cox",
    penalty.factor = pf,
    alpha = 1,
    nfolds = 10
  )

  list(cv = cv_fit, penalized = penalized, forced = forced)
}

fit_m_lasso <- inspect_lasso_fit(heart_male_complete)
fit_f_lasso <- inspect_lasso_fit(heart_female_complete)

plot_lasso_path <- function(lasso_list, sex_label) {
  cv_fit <- lasso_list$cv

  path_mat <- as.matrix(cv_fit$glmnet.fit$beta)
  log_lam <- log(cv_fit$glmnet.fit$lambda)

  path_df <- as.data.frame(t(path_mat)) %>%
    mutate(log_lambda = log_lam) %>%
    pivot_longer(-log_lambda, names_to = "variable", values_to = "coef")
  
  # plot(cv_fit)

  ggplot(path_df, aes(x = log_lambda, y = coef, color = variable)) +
    geom_line(linewidth = 0.8) +
    geom_vline(xintercept = log(cv_fit$lambda.min), linetype = "dashed",
               color = "black", linewidth = 0.8) +
    geom_vline(xintercept = log(cv_fit$lambda.1se), linetype = "dashed",
               color = "red", linewidth = 0.8) +
    annotate("text", x = log(cv_fit$lambda.min), y = Inf,
             label = "lambda.min", hjust = -0.1, vjust = 1.5, size = 3) +
    annotate("text", x = log(cv_fit$lambda.1se), y = Inf,
             label = "lambda.1se", hjust = 1.2, vjust = 1.5, size = 3,
             color = "red") +
    labs(
      title = paste("LASSO Coefficient Path,", sex_label),
      x = "log(lambda)",
      y = "Coefficient",
      color = "Variable"
    ) +
    theme_bw(base_size = 11)
}

plot_lasso_path(fit_m_lasso, "Male")
plot_lasso_path(fit_f_lasso, "Female")


check_lambda <- function(lasso_list, s) {
  coefs <- coef(lasso_list$cv, s = s)
  pen <- lasso_list$penalized
  data.frame(
    variable = pen,
    coef = round(coefs[pen, ], 4),
    selected = coefs[pen, ] != 0
  )
}

# Checking appropriate interim lambdas
# check_lambda(fit_m_lasso, s = exp(-5))
# check_lambda(fit_m_lasso, s = exp(-4))
# check_lambda(fit_f_lasso, s = exp(-6))
# check_lambda(fit_f_lasso, s = exp(-5))

fit_cox <- function(df, lasso_list, chosen_lambda) {
  coefs <- coef(lasso_list$cv, s = chosen_lambda)
  penalized <- lasso_list$penalized
  forced <- lasso_list$forced

  selected_penalized <- penalized[coefs[penalized, ] != 0]
  final_vars <- c(forced, selected_penalized)
  surv_eqn <- as.formula(
    paste("Surv(time_stroke_yrs, censored) ~", paste(final_vars, collapse = " + "))
  )

  cox_fit <- coxph(surv_eqn, data = df, x = TRUE)
  list(cox = cox_fit, selected = selected_penalized, vars = final_vars)
}

# Selected intermediate lambda values that keep variables with some signal
lambda_m <- exp(-4)   
lambda_f <- exp(-5)   

fit_m <- fit_cox(heart_male_complete, fit_m_lasso, lambda_m)
fit_f <- fit_cox(heart_female_complete, fit_f_lasso, lambda_f)

summary(fit_m$cox)
summary(fit_f$cox)
```
# Risk Profiles

```{r}

```


# Questions
-- Clarify time frame: first ten years of follow-up? Yes, 3 periods? Use 365.25 * 10 as 10 years, don't include period 3 (1st period only)
  -- time 0, period 1, stroke 1 = stroke before 
-- Competing risks (comment on limitations but don't account for them)
-- Censor strokes after 10 years
-- exclusion criteria: prior stroke
-- Risk profile: know want to look at age, diabetes, systolic blood pressure; not sure about CHD, blood pressure meds, smoking, total cholesterol, BMI: baseline only 
--Want to keep as much data as possible
--Risk profiles: estimates for 10-year survival for ages 40, 50, and 60 across baseline no risk factors (diabetes, SBP, maybe CHD, BP Meds, smoker, cholesterol, BMI: average continuous, 0's on categoricals including BP meds); high blood pressure (SYS 160+); diabetes; high blood pressure and diabetes; 1 additional profile based on included covariates  
--want survival prob for person of each profile
--More interested in within-sex differences; not necessarily interested in between-sex
--PH assumption: Look at for everything

only descriptive time varying on diabetes and sys BP
2 separate models for M/F (diff covariates, etc.)

Null: HR = 1 for each of age, diabetes, sys BP
367 competing deaths, 111 strokes; standard cox model treats deaths as censored, which overestimates cumulative stroke incidence 

Should be 32 excluded for stroke
Table 1
----
Stroke during follow up
Follow up Time

n = 4275 (127 missing)
no need to exclude high BP & cholesterol

Table 1
Results Table M 
Results Table F
K-M plot (6 panels, age, BP, etc)
Cox PH (6 panels?)
