---
title: "Project 3 Analysis"
author: "Marshall Roll"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(tidyverse)
library(survival)
library(survminer)
```

```{r}
# Raw data stored locally in Documents\Advanced_Methods_Roll\Project 3\Code
heart <- read_csv("../DataRaw/frmgham2.csv")
```

# Data Cleaning
```{r}
# n = 4434
heart <- heart %>% 
  filter(PERIOD == 1) %>% #baseline covariates
  filter(STROKE != 1) %>% #exclude prior stroke # n = 4019
  mutate(time_stroke_yrs = TIMESTRK/365.25) %>% # Convert time stroke to years
  mutate(
    time_stroke_yrs = pmin(time_stroke_yrs, 10),
    censored = ifelse(time_stroke_yrs < 10, 1, 0),
    SEX = case_when(SEX == 1 ~ "Male",
                     SEX == 2 ~ "Female")
  )

heart_male <- heart %>% filter(SEX == "Male") # n = 1751
heart_female <- heart %>% filter(SEX == "Female") # n = 2268
```

# Table 1
```{r}

```

# Overall

```{r}
surv<-survfit(Surv(time_stroke_yrs, censored)~SEX, data=heart)

ggsurvplot(surv, xlim=c(0, 10))
```

# Male

```{r}
coxph(Surv(time_stroke_yrs, censored)~AGE + DIABETES + SYSBP, 
      data = heart_male)
```
# Female

```{r}
coxph(Surv(time_stroke_yrs, censored)~AGE + DIABETES + SYSBP, 
      data = heart_female)
```


# Questions
-- Clarify time frame: first ten years of follow-up? Yes, 3 periods? Use 365.25 * 10 as 10 years, don't include period 3 (1st period only)
  -- time 0, period 1, stroke 1 = stroke before 
-- Competing risks (comment on limitations but don't account for them)
-- Censor strokes after 10 years
-- exclusion criteria: prior stroke
-- Risk profile: know want to look at age, diabetes, systolic blood pressure; not sure about CHD, blood pressure meds, smoking, total cholesterol, BMI: baseline only 

only descriptive time varying on diabetes and sys BP
2 separate models for M/F (diff covariates, etc.)